---
title: "Reproducible Research - Assignment 1"
output: html_document
---

Data Processing
---

* Loading libraries and checking the data

- No of rows should equal 17,568

```{r}
library(lubridate)
library(dplyr)

activity <- read.csv("activity.csv")
head(activity)
tail(activity)
summary(activity)
print(paste("Number of rows",nrow(activity)))
str(activity)
```


Preprocessing the data : converting dates column into dates

```{r}
activity$date <- ymd(activity$date)
str(activity)
```

Analysis Results : Part 1
---
Calculate the mean and median total number of steps taken per day (ignoring missing values)?


```{r}
Tab1 <- filter(activity, !is.na(activity$steps)) %>% group_by(date) %>% summarize(total_step=sum(steps, na.rm=TRUE))
print(paste("Mean: " , mean(Tab1$total_step)))
print(paste("Median: ", median(Tab1$total_step)))
```

Histogram of the Total Number of Steps taken each day

```{r}
hist(Tab1$total_step, col="blue", main="Histogram of Total Number of Steps per Day", xlab = "Total Number of Steps", ylab="No of Days",ylim=c(0,30))
```

Plotting the average daily activity pattern for each 5-minute interval

```{r}
Tab2 <- filter(activity, !is.na(activity$steps)) %>% group_by(interval) %>% summarize(Avg_Steps = mean(steps))
with(Tab2, plot(interval, Avg_Steps, type = "l", xlab = "Interval", ylab = "Average No of Steps", main = "Average No of Steps per Day"))
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
maxstp<-max(Tab2$Avg_Steps)
A <- filter(Tab2, Avg_Steps == maxstp)
print(paste("5 minute interval with the maximum number of steps", A[,1]))
```

Imputing missing values
---

#### Calculating the number of days/intervals where there are missing values (coded as NA ) since the presence of missing days may introduce bias into some calculations or summaries of the data.

- Calculating the total number of missing values in the dataset (i.e. the total number of rows with  NA s)

```{r}
print(paste("Number of rows with NAs = ", sum(is.na(activity$steps))))
```

- Imputing the missing values by using the daily Average Steps for the missing values.

- Select Non-missing obs into a object - label as Measured

```{r}
Tab2_Act <- filter(activity, !is.na(activity$steps))
Tab2_Act$partition <- 'Measured'
```

- Select Missing obs into a object - label as Imputed

```{r}
Tab2_NA <- filter(activity, is.na(activity$steps))
Tab2_NA$partition <- 'Imputed'
```
- Merge the Interval Averages onto the NA dataset and create a new dataset with both the measured and imputed values

```{r}
Tab2_ordered <- merge(Tab2_NA, Tab2, by.x = "interval", by.y = "interval")
Tab2_ordered$steps <- Tab2_ordered$Avg_Steps

Tab3 <- rbind(Tab2_Act[,1:4], Tab2_ordered[,1:4])
```

- Check the rows after the merge and appending

```{r}
print(paste("Number of rows = ",nrow(Tab3)))
```

- Calculating the Mean and Median of this dataset.
- Histogram of the Total Number of Steps each Day using the dataset including the Imputed values.

```{r}
Tab4 <- filter(Tab3, !is.na(activity$steps)) %>% group_by(date) %>% summarize(total_step=sum(steps, na.rm=TRUE))
print(paste("Mean: " , mean(Tab4$total_step)))
print(paste("Median: ", median(Tab4$total_step)))

hist(Tab4$total_step, col="blue", main="Histogram of Total Number of Steps per Day", xlab = "Total Number of Steps", ylab="No of Days",ylim=c(0,30))

print(paste("Difference in Mean (Measured only - incl Imputed) = ", mean(Tab4$total_step) - mean(Tab1$total_step)))
print(paste("Difference in Median (Measured only - incl Imputed) = ", median(Tab4$total_step) - median(Tab1$total_step)))
```

* Differences in the mean and median values between the dataset including the imputed values and the measured dataset is 320 and -211 respectively. The impact of imputing missing data is that the mean and median values can change away from those of the 'measured' dataset and this could possibly introduce unexpected bias into the results.

Analysis Results : Part 2 - Activity between weekdays and weekends.
---

For this part of the analysis the dataset including the imputed values was used.

- Including a new factor variable with two levels -- "weekday" and "weekend" which indicates whether a given date is a weekday or weekend day.
```{r}
Tab3$weekdays <- weekdays(Tab3$date)
Tab3$TypeofDay <- Tab3[]

Tab3$TypeofDay <- with(Tab3,ifelse(Tab3$weekdays %in% c("Saturday","Sunday"),"weekend","weekday"))
```

- Comparing the average number of steps taken over week or weekend days by using a time series plot of the 5-minute interval (x-axis). 

```{r}
Tab5 <- Tab3 %>% group_by(TypeofDay, interval) %>% summarize(Avg_Steps = mean(steps))

library(lattice)
xyplot(Avg_Steps ~ interval | TypeofDay, data = Tab5, layout = c(1,2), type = 'l')
```

* On average the weekdays has a higher step rate up to approx 9:30 after which there is a fairly constant average step rate up to 20:00. On weekends there is less variation in the average step rate for most of the day. 

